Mission-time LTL (MLTL) Formula Validation Via Regular Expressions\thanks{Work supported in part by NSF grant DMS-1950583}

### 2023-06-15 paper/Paper.tex
updated paper

\section{WEST Algorithm and Analysis} \label{WESTalg}
We now present the  full WEST algorithm, which recursively computes all satisfying temporal regular expressions to a given MLTL formula (in NNF). 
For convenience, we use sets to represent alternation of regular expressions; for $n$ regular expressions $t_0, ..., t_{n-1}$, we'll write $\{t_0, ..., t_{n-1}\} = \bigcup_{i=0}^{n-1} t_i$. 
Additionally, we provide details for performing set intersection of temporal regular expressions, and the algorithm for the temporal operator $\mathcal{U}$. 
Note how \texttt{reg\_U} parallels the regular expression defined for $\mathcal{U}$, and the algorithms for the other three temporal operators follow an identical structure.

One major detail that drastically improves running time is applying a simplifying function to the output of set intersection. 
This is because the set intersection operation is repeatedly used to compute the temporal operators $\mathcal{G}, \mathcal{U}, \mathcal{R}$. 
Although repeated set intersection blows up exponentially in space in the worse case, it rarely does in practice; thus why we frequently simplify outputs. 
We simplify alternation by using the fact that for regular expressions $w_0$ and $w_1$, $\mathscr{L}(w_0 1 w_1 \lor w_0 0 w_1) = \mathscr{L}(w_0 s w_1)$, and check if the reduction applies at every index between every pair of strings in a given set.

@@

% ZILI: Presenting algorithm vertically instead
\hspace{-5mm}
\begin{minipage}{0.52\textwidth} 
\begin{algorithm}[H] 
\caption{WEST Algorithm}
Inputs: $\phi$ - MLTL formula in NNF \\
$\phi_1$ and $\phi_2$ below are subformulas of $\phi$\\
$n$ - number of propositional variables \\
Output: set of REGEX satisfying $\phi$
\begin{algorithmic}[1]
    \Procedure{reg}{string $\phi$, int $n$}
        \If{$\phi$ is $\top$ or $\bot$}
            \State return reg\_prop\_const($\phi$, n)
        \EndIf

        \If{$\phi = \phi_1 \mathcal{R}_{[a, b]} \phi_2$}
            \State return reg\_R(reg($\phi_1$), reg($\phi_2$), a, b, n)
        \EndIf
    \EndProcedure
\end{algorithmic}
\end{algorithm}
\vspace{5.9mm}
\end{minipage}
\hfill
\begin{minipage}{0.48\textwidth}
\begin{algorithm}[H]
\caption{set\_intersect}
Inputs: $R, T$ - two sets of REGEX\\
$n$ - number of propositional variables\\
Output: set of REGEX equal to $R \land T$
\begin{algorithmic}[1]
\Procedure{set\_intersect}{$R$, $T$, n}
    \State Pad($R$, $T$, n), return $\leftarrow \{\}$
    \For{$(r, t) \in R\times T$}
            \State add bit\_wise\_and($r, t$) to ret 
    \EndFor
    \State return simplify(ret)
\EndProcedure
\end{algorithmic}
\end{algorithm}
\vspace{-6.5mm}
\begin{algorithm}[H]
\caption{reg\_U} 
Inputs: $r_\phi$, $r_\psi$ - sets of REGEX for MLTL formulas $\phi$ and $\psi$ (after calling reg)\\
$a, b$ - integers representing interval bound \\
$n$ - number of propositional variables\\
Output: set of REGEX for $\phi \mathcal{U}_{[a, b]} \psi$
\begin{algorithmic}[1]
\Procedure{reg\_U}{$r_\phi$, $r_\psi$, $a$, $b$, $n$} 
    \State comp $\leftarrow$ ((`$S$')$^{n}$ + `,')$^{a} + $ r\_$\psi$
    \If{$a > b$} return $\{\}$
    \EndIf
    \For{($a \leq i \leq b-1$) }
        \State G1 $\leftarrow$ reg\_G($r_\phi$, $a$, $i$, $n$)
        \State \parbox[t]{.8\linewidth}{G2 $\leftarrow$ reg\_G($r_\psi$, $i+1$, $i+1$, $n$)}
        \State \parbox[t]{.8\linewidth}{temp\_comp $\leftarrow$ set\_intersect(G1,  \hspace*{5mm}G2, n)}
        \State \parbox[t]{.8\linewidth}{comp $\leftarrow$ join( comp, temp\_comp)}
        % , \hspace*{5mm}simp)}
    \EndFor
    \State return comp
\EndProcedure
\end{algorithmic}
\end{algorithm}
\end{minipage}

@@

\vspace{-3mm}$$\mathscr{L}(R) \cap \mathscr{L}(T) = \left(\bigcup_{i = 0}^{a-1} \mathscr{L}(r_i) \right)\cap \left( \bigcup_{j = 0}^{b-1} \mathscr{L}(t_j)\right) = \bigcup_{i = 0}^{a-1} \bigcup_{j = 0}^{b-1} \left(\mathscr{L}(r_i) \cap \mathscr{L}(t_j)\right).$$

@@

    \vspace{-.4cm}

@@

    where $\mathcal{C}_U = \delta ((n+1)(b-1) + 1)^\delta \cdot (n+1) \cdot d$.\\

@@

    \vspace{-0.2cm}

@@

For time complexity, through a similar analysis, we've found that the time-complexity of all of the above functions is the same as their space complexity. In particular, the time complexities of $\phi = p_k, \scriptstyle{\sim}\textstyle p_k, \top, \bot$, as well as that of \texttt{set\_intersect}, \texttt{join}, and the operator $\mathcal{U}$ match their respective space complexities.
\begin{theorem}[\emph{Time Complexity}]
Given a well formed MLTL formula $\phi$, $\emph{\texttt{reg}}(\phi)$ has worst-case time complexity that is $O(\mathcal{C}_U^{\delta ^ \ell} \cdot (\ell + 1)^{\delta^{\ell + 1}})$, where $\ell$ is the number of logical connectives ($\land, \lor, \mathcal{F}, \mathcal{G}, \mathcal{R}, \mathcal{U}$) in $\phi$. 
\end{theorem}
If in the worst case no simplification occurs in any call of \texttt{set\_intersect}, space complexity remains unchanged, but simplifying a set of regular expressions is cubic in input size. In practice however, both time and space complexities are much more optimistic than worst-case estimates.

\subsection{Experimental Benchmarking}
We accompany theoretical space and time complexity with experimental evaluation of these complexities using randomly-generated MLTL formulas. What we observed from the simulations is that the worst case complexity, both for space and time, is relatively rare. We also observe that these worst cases are extreme outliers and that in nearly every case, are examples of nested binary temporal operators. As seen in \cite{HLR21}, \cite{KZJZR20}, \cite{AJR22}, and \cite{HCHJR21}, nested binary temporal operators do not appear in any specifications, and thus are likely rare in the literature and in practical applications.

We ran these experiments on an Intel(R) Core(TM) i7-4770S CPU at 3.10GHz with 32gb RAM.
For each simulation, we generated 1000 MLTL formulas using the parameters \texttt{delta}, \texttt{interval\_max}, number of propositional variables, and number of iterations. \texttt{delta} is the maximum length we allow for any interval. \texttt{interval\_max} is the largest allowed upper bound for any interval. Number of iterations refers to the level of nesting in the generated formulas. For example, $\mathcal{G}_{[0,2]}p_0$ is a formula generated with one iteration, while $\mathcal{G}_{[0,2]}(p_0 \land p_1)$ is a formula generated with two iterations. We measure the number of characters in the output and the amount of time in microseconds taken to run the program. Note that these simulations run the \texttt{reg} function with the subformula boolean set to false and the simplify boolean set to true.\\
\noindent\textbf{Simulation 1}
For the first simulation, we consider 2 iterations, 5 propositional variables, 10 for \texttt{delta}, and 10 for \texttt{interval\_max}. We obtain plots \ref{InVsOutputLen} and \ref{InLenVsRunTime}:\\
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim1Length_large.png}
    \captionof{figure}{Input length vs output length.\vspace{3mm}}
    \label{InVsOutputLen}
\end{minipage}
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim1Time_large.png}
    \captionof{figure}{Input length vs run time.\vspace{3mm}}
    \label{InLenVsRunTime}
\end{minipage}\\
\noindent The formula $(p_0 = p_1) \mathcal{U}_{[2,9]} (p_1 \mathcal{U}_{[7,9]} p_3)$ was an outlier for both Figures \ref{InVsOutputLen} and \ref{InLenVsRunTime}, and the formulas $(p_4 \rightarrow p_2) \mathcal{R}_{[1,8]} (p_3 \mathcal{U}_{[3,4]} p_0)$ and $(p_3 \mathcal{R}_{[2,7]} p_4) \mathcal{R}_{[1,9]} (p_4 \mathcal{R}_{[4,9]} p_3)$ were outliers for Figure \ref{InLenVsRunTime}. \\
\noindent\textbf{Simulation 2}
For the second simulation, we consider 1 iteration, 10 propositional variables, 20 for \texttt{delta}, and 20 for \texttt{interval\_max}. Here are some examples of generated formulas:
$\mathcal{F}_{[10,17]} p_4$, $\neg p_6$, $p_6 \mathcal{U}_{[6,14]} p_8$. We obtain plots \ref{fig:InVsOutLen2} and \ref{fig:InLenVsRunTime2}:\\
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim2Length_large.png}
    \captionof{figure}{Input length vs output length\vspace{3mm}}
    \label{fig:InVsOutLen2}
\end{minipage}
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim2Time_large.png}
    \captionof{figure}{Input length vs run time.\vspace{3mm}}
    \label{fig:InLenVsRunTime2}
\end{minipage}\\
\noindent\textbf{Simulation 3}
For the third simulation, we consider 2 iterations, 10 propositional variables, 5 for \texttt{delta}, and 10 for \texttt{interval\_max}. We obtain plots \ref{fig:InLenVsOutLen3} and \ref{fig:InLenVsRunTime3}:\\
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim3Length_large.png}
    \captionof{figure}{Input length vs output length\vspace{3mm}}
    \label{fig:InLenVsOutLen3}
\end{minipage}
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim3Time_large.png}
    \captionof{figure}{Input length vs run time.\vspace{3mm}}
    \label{fig:InLenVsRunTime3}
\end{minipage}\\
The formulas $(p_7 \mathcal{U}_{[5,7]} p_9) \mathcal{U}_{[3,7]} (\mathcal{F}_{[1,4]} p_4)$ and $\mathcal{G}_{[3,7]} (p_9 \mathcal{U}_{[0,4]} p_0)$ were outliers in both Figures \ref{fig:InLenVsOutLen3} and \ref{fig:InLenVsRunTime3}, and Figure \ref{fig:InLenVsOutLen3} had one more outlier from the formula $(\neg p_9) \mathcal{R}_{[5,9]} (p_5 \mathcal{U}_{[2,4]} p_3)$.
\noindent\textbf{Simulation 4}
For the fourth simulation, we consider 1 iteration, 5 propositional variables, 10 for \texttt{delta}, and 10 for \texttt{interval\_max}. Some examples of generated formulas are: $\neg p_1$, $p_1 \mathcal{R}_{[6,7]} p_3$, $\mathcal{F}_{[5,5]} p_3$. We obtain plots \ref{InLenVsOutLen4} and \ref{InLenVsRunTime4}:\\
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim4Length_large.png}
    \captionof{figure}{Input length vs output length\vspace{3mm}}
    \label{InLenVsOutLen4}
\end{minipage}%
\begin{minipage}{0.5 \textwidth}
    \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{images/Sim4Time_large.png}
    \captionof{figure}{Input length vs run time.\vspace{3mm}}
    \label{InLenVsRunTime4}
\end{minipage}
What we conclude from the simulations is that for most practical purposes, the WEST algorithm demonstrates good space and time complexity.
\section{Correctness of WEST Tool Implementation} \label{test}
We accompany our proof of algorithmic correctness with a rigorous evaluation of implementation correctness, showing that our WEST tool correctly implements our WEST algorithm. Since our proof of algorithmic correctness is manual and our focus is on usability for validation by humans, we utilize more traditional techniques for robust software engineering with testing-based evaluation. An attractive alternative for future work involves encoding our proofs in an interactive theorem prover and utilizing a verified code generator like PVS2C\cite{Sha17} or the Isabelle/HOL code generator \cite{HB21}. 
Since the primary purpose of the WEST tool is validation of formulas for humans, we focused on writing a tool with an intuitive human-facing interface and then verifying that implementation rather than wrestling with verified code generated by a tool made to optimize for different objectives.\\
One na\"ive approach involves testing every possible input up to a certain size to verify the output or behaviors of the program.
For WEST, this strategy would generate a very large test suite with significant redundancy. 
For instance, there is little sense in testing all MLTL formulas of the form $p_0 \mathcal{U}_{[0,t]} p_1$ for all $t$ such that $0 \leq t \leq 99$; verification of a few should give sufficient confidence of correctness of the program. 
We approach this problem more intelligently: we test our implementation with a test suite that explores all control paths (up to a certain depth) of the algorithm. A control path is a sequence of lines of code that are executed in one calculation of an input formula. Differing structures of formulas will lead to different sequences of lines of code to be executed, and a test suite to explore all such paths provides high confidence in the correctness of the WEST tool.

\subsection{Intelligent Fuzzing}
Negative testing is the process of checking that a program is able to handle unexpected inputs without crashing; we employ fuzz testing (also know as fuzzing), a form of negative testing. We deviate from traditional black-box fuzzing defined by \cite{Miller}:
\textit{``If we consider a program to
be a complex finite state machine, then our testing strategy can be thought of as a random walk through the state space, searching for undefined states.''}\\
Instead we utilize intelligent fuzzing (sometimes referred to as whitebox fuzzing), which is an alternate approach to fuzz testing that leverages knowledge about program structure to generate valid inputs and increase coverage. 
Borrowing the words of Miller, our approach to testing the WEST program can be thought of as walking all possible paths up to a certain depth of the state space of our algorithm. 
We first outline our overall approach to intelligent fuzzing: 
\begin{enumerate}
    \item Construct a state diagram of our algorithm, represented as a directed graph. 
    The edges of the graph capture control flow of our algorithm, and vertices represent the remaining blocks of the code without branching statements.  
    \item Construct a test suite that explores all possible paths in the digraph up to a certain depth. Run the WEST program on the test suite to produce a set of output files.
    \item Run a na\"ive brute force generator of satisfying computations on the test suite and verify that both outputs match for all test cases.
\end{enumerate}
\subsubsection{State Diagram Construction}
 We can represent the state space of the WEST algorithm as a directed graph with the edges representing the control flow and vertices representing blocks of contiguous code without branching statements. 
 The core of the WEST program lies in the recursive routine, \texttt{reg}, which calls the $8$ different subroutines as shown in \ref{fig:abstracted_graph}.
\begin{figure}[t]
    \centering
    \begin{tikzpicture}[ node/.style={rounded rectangle, draw=black!60, fill=black!5, thick, minimum size=8mm},
    bignode/.style={rounded rectangle, draw=black!60, fill=black!5, thick, minimum size=10mm},
    star/.style={circle, draw=red!60, fill=red!5, thick, minimum size=1mm} ]

@@

For $d = 0$, only the paths leading to \texttt{prop\_var} and \texttt{prop\_cons} can be explored, so $\phi(0) = 2$. For $d \geq 1$, we recursively calculate $\phi(d + 1) = 2\phi(d) + 4\phi(d)^2$; paths to \texttt{reg\_G} and \texttt{reg\_F} is counted in the linear term, and paths to \texttt{reg\_U}, \texttt{reg\_R}, \texttt{set\_intersect}, and \texttt{join} is counted by the quadratic term. 

@@

The full implementation details are available in the WEST GitHub repository.\footnote{Links have been redacted for blind review.} 

@@

In comparison, the WEST program executed the same test suite in under thirty minutes on the same machine. Note that the brute force program outputs only computations of zeros and ones, and thus comparing the outputs of the WEST program requires expanding out the ``$S$'' characters in the regular expressions. For instance, the regular expression ``$1SS0SS1$'' yields $2^{\text{\# of `$S$'s}} = 2^4 = 16$ computations after expansion.

@@

\section{Regular Expression Simplification Theorem} \label{simpsection}
As a final result, we provide a regular expression simplification theorem. This theorem describes a form of a set of computations that are guaranteed to simplify down to a string of all ``$S$'' characters. Although not implemented in the WEST program, this theorem may help users identify tautologies, as the WEST program does not always output a single string of ``$S$'' characters when a formula holds true for every computation. We first define some vocabulary.

>>>>

\section{WEST Algorithm and Analysis} \label{WESTalg}
Algorithm 1 recursively computes all satisfying temporal regular expressions to an input formula $\phi$.
We use sets to represent alternation of regular expressions; for $n$ regular expressions $t_0, ..., t_{n-1}$, we'll write $\{t_0, ..., t_{n-1}\} = \bigcup_{i=0}^{n-1} t_i$. 
Additionally, we provide details for performing set intersection of temporal regular expressions, and the algorithm for the temporal operator $\mathcal{U}$.
Note how \texttt{reg\_U} parallels the regular expression defined for $\mathcal{U}$, and the algorithms (see appendix \ref{pseudocode_appendix}) for the other three temporal operators follow an identical structure.

For for regular expressions $w_0$ and $w_1$, a useful reduction is that $\{w_0 1 w_1, w_0 0 w_1\} = \{w_0 s w_1\}$. Each time set intersection is performed, we greedily apply this reduction to all appropriate pairs of regular expressions in the set. This prevents repeated set intersection operations from blowing up exponentially most of the time, and drastically improves running time.

% Applying a simplifying function to the output of set intersection drastically improves running time. 
% This is because the set intersection operation is repeatedly used to compute the temporal operators $\mathcal{G}, \mathcal{U}, \mathcal{R}$. 
% Although repeated set intersection blows up exponentially in space in the worse case, it rarely does in practice; thus why we frequently simplify outputs. 
% We simplify alternation by using the fact that for regular expressions $w_0$ and $w_1$, $\mathscr{L}(w_0 1 w_1 \lor w_0 0 w_1) = \mathscr{L}(w_0 s w_1)$, and check if the reduction applies at every index between every pair of strings in a given set.

@@

% % ZILI: Presenting algorithm vertically instead
% \hspace{-5mm}
% \begin{minipage}[t]{0.52\textwidth} 
% \begin{algorithm}[H] 
% \caption{WEST Algorithm}
% Inputs: $\phi$ - MLTL formula in NNF \\
% $\phi_1$ and $\phi_2$ below are subformulas of $\phi$\\
% $n$ - number of propositional variables \\
% Output: set of REGEX satisfying $\phi$
% \begin{algorithmic}[1]
%     \Procedure{reg}{string $\phi$, int $n$}
%         \If{$\phi$ is $\top$ or $\bot$}
%             \State return reg\_prop\_const($\phi$, n)
%         \EndIf

%         \If{$\phi$ is $p_k$ or $\neg p_k$}
%             \State return reg\_prop\_var($\phi$, n)
%         \EndIf

%         \If{$\phi  = \phi_1 \land \phi_2$}
%             \State \parbox[t]{.82\linewidth}{return set\_intersect(reg($\phi_1$), reg($\phi_2$), \hspace*{5mm}n)}
%         \EndIf

%         \If{$\phi  = \phi_1 \lor \phi_2$}
%             \State return join(reg($\phi_1$), reg($\phi_2$), n)
%         \EndIf

%         \If{$\phi = \mathcal{F}_{[a, b]}\phi_1$}
%             \State return reg\_F(reg($\phi_1$), a, b, n)
%         \EndIf

%         \If{$\phi = \mathcal{G}_{[a, b]}\phi_1$}
%             \State return reg\_G(reg($\phi_1$), a, b, n)
%         \EndIf

%         \If{$\phi = \phi_1 \mathcal{U}_{[a, b]} \phi_2$}
%             \State return reg\_U(reg($\phi_1$), reg($\phi_2$), a, b, n)
%         \EndIf

%         \If{$\phi = \phi_1 \mathcal{R}_{[a, b]} \phi_2$}
%             \State return reg\_R(reg($\phi_1$), reg($\phi_2$), a, b, n)
%         \EndIf
%     \EndProcedure
% \end{algorithmic}
% \end{algorithm}
% \vspace{5.9mm}
% \end{minipage}
% \hfill
% \begin{minipage}[t]{0.48\textwidth}
% \begin{algorithm}[H]
% \caption{set\_intersect}
% Inputs: $R, T$ - two sets of REGEX\\
% $n$ - number of propositional variables\\
% Output: set of REGEX equal to $R \land T$
% \begin{algorithmic}[1]
% \Procedure{set\_intersect}{$R$, $T$, n}
%     \State Pad($R$, $T$, n), return $\leftarrow \{\}$
%     \For{$(r, t) \in R\times T$}
%             \State add bit\_wise\_and($r, t$) to ret 
%     \EndFor
%     \State return simplify(ret)
% \EndProcedure
% \end{algorithmic}
% \end{algorithm}
% \vspace{-6.5mm}
% \begin{algorithm}[H]
% \caption{reg\_U} 
% Inputs: $r_\phi$, $r_\psi$ - sets of REGEX for MLTL formulas $\phi$ and $\psi$ (after calling reg)\\
% $a, b$ - integers representing interval bound \\
% $n$ - number of propositional variables\\
% Output: set of REGEX for $\phi \mathcal{U}_{[a, b]} \psi$
% \begin{algorithmic}[1]
% \Procedure{reg\_U}{$r_\phi$, $r_\psi$, $a$, $b$, $n$} 
%     \State comp $\leftarrow$ ((`$S$')$^{n}$ + `,')$^{a} + $ r\_$\psi$
%     \If{$a > b$} return $\{\}$
%     \EndIf
%     \For{($a \leq i \leq b-1$) }
%         \State G1 $\leftarrow$ reg\_G($r_\phi$, $a$, $i$, $n$)
%         \State \parbox[t]{.8\linewidth}{G2 $\leftarrow$ reg\_G($r_\psi$, $i+1$, $i+1$, $n$)}
%         \State \parbox[t]{.8\linewidth}{temp\_comp $\leftarrow$ set\_intersect(G1,  \hspace*{5mm}G2, n)}
%         \State \parbox[t]{.8\linewidth}{comp $\leftarrow$ join( comp, temp\_comp)}
%         % , \hspace*{5mm}simp)}
%     \EndFor
%     \State return comp
% \EndProcedure
% \end{algorithmic}
% \end{algorithm}
% \end{minipage}

@@

%\vspace{-3mm}
$$\mathscr{L}(R) \cap \mathscr{L}(T) = \left(\bigcup_{i = 0}^{a-1} \mathscr{L}(r_i) \right)\cap \left( \bigcup_{j = 0}^{b-1} \mathscr{L}(t_j)\right) = \bigcup_{i = 0}^{a-1} \bigcup_{j = 0}^{b-1} \left(\mathscr{L}(r_i) \cap \mathscr{L}(t_j)\right).$$

@@

    {\setlength{\abovedisplayskip}{0pt}\setlength{\belowdisplayskip}{0pt}

@@

    \end{align*}}
    where $\mathcal{C}_U = \delta ((n+1)(d-1) + 1)^\delta \cdot (n+1) \cdot d$.\\

@@

    % \vspace{-0.2cm}

@@

Through a similar analysis we've found that the time complexity of all of the above functions is the same as their space complexity.
\begin{theorem}[\emph{Time Complexity}]
Given a well formed MLTL formula $\phi$, $\emph{\texttt{reg}}(\phi)$ has worst-case time complexity that is $O(\mathcal{C}_U^{\delta ^ \ell} \cdot (\ell + 1)^{\delta^{\ell + 1}})$, where $\ell$ is the number of logical connectives ($\land, \lor, \mathcal{F}, \mathcal{G}, \mathcal{R}, \mathcal{U}$) in $\phi$. 
If in the worst case no simplification occurs in any call of \texttt{set\_intersect}, space complexity remains unchanged, but simplifying a set of regular expressions is cubic in input size. In practice however, both time and space complexities are much more optimistic than worst-case estimates.

\subsection{Experimental Benchmarking} \label{WESTexp}
We accompany theoretical space and time complexity with experimental evaluation of these complexities using randomly-generated MLTL formulas. What we observed from the simulations is that the worst case complexity, both for space and time, is relatively rare. We also observe that these worst cases are extreme outliers and that in nearly every case, are examples of nested binary temporal operators. As seen in \cite{HLR21}, \cite{KZJZR20}, \cite{AJR22}, and \cite{HCHJR21}, nested binary temporal operators do not appear in any specifications, and thus are likely rare in the literature and in practical applications.

We ran these experiments on an Intel(R) Xeon(R) Gold 6140 CPU @ 2.30GHz with 376 GB RAM.
For each simulation, we generated 1000 MLTL formulas using the parameters \texttt{delta}, \texttt{interval\_max}, number of propositional variables, and number of iterations. \texttt{delta} is the maximum length we allow for any interval. \texttt{interval\_max} is the largest allowed upper bound for any interval. Number of iterations refers to the level of nesting in the generated formulas. For example, $\mathcal{G}_{[0,2]}p_0$ is a formula generated with one iteration, while $\mathcal{G}_{[0,2]}(p_0 \land p_1)$ is a formula generated with two iterations. We measure the number of characters in the output and the amount of time in microseconds taken to run the program. Note that these simulations run the \texttt{reg} function with the subformula boolean set to false and the simplify boolean set to true.\\
\noindent\textbf{Simulation 1}
For the first simulation, we consider 2 iterations, 5 propositional variables, 10 for \texttt{delta}, and 10 for \texttt{interval\_max}. We obtain plots \ref{InVsOutputLen1} and \ref{InLenVsRunTime1}.
\begin{figure}[H]
\captionsetup[subfigure]{aboveskip=-1pt,belowskip=-1pt}
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTspace1.png}
    \caption{}
    \label{InVsOutputLen1}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTtime1.png}
    \caption{}
    \label{InLenVsRunTime1}
    \end{subfigure}
    \caption{$(p_0 = p_1) \mathcal{U}_{[2,9]} (p_1 \mathcal{U}_{[7,9]} p_3)$ is an outlier in both. $(p_4 \rightarrow p_2) \mathcal{R}_{[1,8]} (p_3 \mathcal{U}_{[3,4]} p_0)$ and
    $(p_3 \mathcal{R}_{[2,7]} p_4) \mathcal{R}_{[1,9]} (p_4 \mathcal{R}_{[4,9]} p_3)$ are outliers only in \subref{InLenVsRunTime1}.}
\end{figure}

\noindent\textbf{Simulation 2}
For the second simulation, we consider 1 iteration, 10 propositional variables, 20 for \texttt{delta}, and 20 for \texttt{interval\_max}. We obtain plots \ref{fig:InVsOutLen2} and \ref{fig:InLenVsRunTime2}.
\begin{figure}[h]
\captionsetup[subfigure]{aboveskip=-1pt,belowskip=-1pt}
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTspace2.png}
    \caption{}
    \label{fig:InVsOutLen2}
    \end{subfigure}
\begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTtime2.png}
    \caption{}
    \label{fig:InLenVsRunTime2}
    \end{subfigure}
    \caption{We observe no outliers.}
\end{figure}

\noindent\textbf{Simulation 3}
For the third simulation, we consider 2 iterations, 10 propositional variables, 5 for \texttt{delta}, and 10 for \texttt{interval\_max}. We obtain plots \ref{fig:InLenVsOutLen3} and \ref{fig:InLenVsRunTime3}.
\begin{figure}[h]
\captionsetup[subfigure]{aboveskip=-1pt,belowskip=-1pt}
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTspace3.png}
    \caption{}
    \label{fig:InLenVsOutLen3}
    \end{subfigure}
    \begin{subfigure}[h]{0.49 \textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTtime3.png}
    \caption{}
    \label{fig:InLenVsRunTime3}
    \end{subfigure}
    \caption{$(p_7 \mathcal{U}_{[5,7]} p_9) \mathcal{U}_{[3,7]} (\mathcal{F}_{[1,4]} p_4)$ and
    $\mathcal{G}_{[3,7]} (p_9 \mathcal{U}_{[0,4]} p_0)$ are outliers in both. $\mathcal{G}_{[3,7]} (p_9 \mathcal{U}_{[0,4]} p_0)$ is an outlier in \ref{fig:InLenVsOutLen3} only.}
\end{figure}

\noindent\textbf{Simulation 4}
For the fourth simulation, we consider 1 iteration, 5 propositional variables, 10 for \texttt{delta}, and 10 for \texttt{interval\_max}. We obtain plots \ref{InLenVsOutLen4} and \ref{InLenVsRunTime4}.
\begin{figure}[h]
\captionsetup[subfigure]{aboveskip=-1pt,belowskip=-1pt}
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTspace4.png}
    \caption{}
    \label{InLenVsOutLen4}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/WESTtime4.png}
    \caption{}
    \label{InLenVsRunTime4}
    \end{subfigure}
    \caption{We observe no outliers.}
\end{figure}

What we conclude from the simulations is that for most practical purposes, the WEST algorithm demonstrates good space and time complexity.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correctness of WEST Tool Implementation} \label{test}
We accompany our proof of algorithmic correctness with a rigorous evaluation of implementation correctness, showing that our WEST tool correctly implements our WEST algorithm. Since our proof of algorithmic correctness is manual and our focus is on usability for validation by humans, we utilize more traditional techniques for robust software engineering with testing-based evaluation. 
% An attractive alternative for future work involves encoding our proofs in an interactive theorem prover and utilizing a verified code generator like PVS2C\cite{Sha17} or the Isabelle/HOL code generator \cite{HB21}. 
% Since the primary purpose of the WEST tool is validation of formulas for humans, we focused on writing a tool with an intuitive human-facing interface and then verifying that implementation rather than wrestling with verified code generated by a tool made to optimize for different objectives.\\
The na\"ive approach is to test all inputs up to a certain size and verify the outputs, but this strategy would generate an unnecessarily large and redundant test suite. 
For instance, there is little sense in testing all MLTL formulas of the form $p_0 \mathcal{U}_{[0,t]} p_1$ for all $t$ such that $0 \leq t \leq 99$; verification of a few should give sufficient confidence of correctness of the program. 
Instead, we test our implementation with a test suite that explores all possible sequences of lines of code that are executed (up to a certain depth).
% control paths (up to a certain depth) of the algorithm. A control path is a sequence of lines of code that are executed in one calculation of an input formula. Differing structures of formulas will lead to different sequences of lines of code to be executed, and a test suite to explore all such paths provides high confidence in the correctness of the WEST tool.

\subsection{Intelligent Fuzzing}
% Negative testing is the process of checking that a program is able to handle unexpected inputs without crashing; we employ fuzz testing (also know as fuzzing), a form of negative testing. We deviate from 
Traditional black-box fuzzing is defined by Miller \cite{Miller}:
\textit{``If we consider a program to
be a complex finite state machine, then our testing strategy can be thought of as a random walk through the state space, searching for undefined states.''}\\
Instead we utilize intelligent fuzzing, an alternate approach that leverages knowledge about program structure to generate valid inputs and increase coverage. 
Borrowing the words of Miller, our approach to testing the WEST program can be thought of as walking all possible paths up to a certain depth of the state space of our algorithm. 
We first outline our overall approach to intelligent fuzzing: 
\begin{enumerate}
    \item Construct a directed graph representation of our algorithm.
    The edges capture control flow of our algorithm, and vertices represent non-branching blocks of code.
    \item Construct a test suite that explores all possible paths in the digraph up to a certain depth. Run the WEST program on the test suite to produce a set of output files.
    \item Run a na\"ive brute force generator of satisfying computations on the test suite and verify that both outputs match for all test cases.
\end{enumerate}
\subsubsection{State Diagram Construction}
 We can represent the state space of the WEST algorithm as a directed graph with the edges representing the control flow and vertices representing blocks of contiguous code without branching statements. 
 The core of the WEST program lies in the recursive routine, \texttt{reg}, which calls the $8$ different subroutines as shown in \ref{fig:abstracted_graph}.
\begin{figure}[h]
    \centering
    \begin{tikzpicture}[ node/.style={rounded rectangle, draw=black!60, fill=black!5, thick, minimum size=8mm},
    bignode/.style={rounded rectangle, draw=black!60, fill=black!5, thick, minimum size=10mm},
    star/.style={circle, draw=red!60, fill=red!5, thick, minimum size=1mm} ]

@@

For $d = 0$, only the paths leading to \texttt{prop\_var} and \texttt{prop\_cons} can be explored, so $\phi(0) = 2$. For $d \geq 1$, we recursively calculate $\phi(d + 1) = 2\phi(d) + 4\phi(d)^2$; paths to \texttt{reg\_G} and \texttt{reg\_F} is counted by the linear term, and paths to \texttt{reg\_U}, \texttt{reg\_R}, \texttt{set\_intersect}, and \texttt{join} is counted by the quadratic term. 

@@

The full implementation details are available in the WEST GitHub repository \href{https://github.com/zwang271/2022-Iowa-State-REU-Temporal-Logic-}{[Link]}.

@@

In comparison, the WEST program executed the same test suite in under thirty minutes on the same machine. Note that the brute force program outputs only computations of zeros and ones, and thus comparing the outputs of the WEST program requires expanding out the ``$S$'' characters in the regular expressions. %For instance, the regular expression ``$1SS0SS1$'' yields $2^{\text{\# of `$S$'s}} = 2^4 = 16$ computations after expansion.

@@

\section{Regular Expression Simplification Theorem} \label{simpsection}
As a final result, we provide a regular expression simplification theorem. This theorem describes the form of a set of computations that simplify down to all ``$S$'' characters. This theorem may help users identify tautologies, as the WEST program does not always output a single string of ``$S$'' characters when a formula holds true for every computation. We first define some vocabulary.
 We say a regular expression composed entirely of ``$S$'' characters and commas is an \emph{arbitrary computation}. For the purposes of the following theorem, we remove all commas from computations. We say a ``0'' or a ``1'' in a regular expression is a \emph{fixed truth value}. We overload the definition of a matrix and say that a \emph{matrix} is a representation of a union of regular expressions, where each row is a regular expression. This aids significantly in the description and proof of the theorem; note, however, that matrix algebra does not apply in this definition.
